---
title: "Week 6: Transformations, Categories, Interactions, & Fixed Effects"
description: "Changing the shape of predictors and testing whether those changes improve prediction."
date: 2026-10-09
---

## What We Did

We extended linear regression by changing the form of the variables given to
the model. We used logs for skewed variables, squared terms for curved
relationships, categories for meaningful groups, interactions for slopes that
vary by group, and fixed effects for differences between places.

For every change, the main prediction question stayed the same: did it lower
cross-validation RMSE? We used the same observations and folds for every model
so that differences in RMSE came from the features rather than from a different
sample or a lucky train/test split.

## Key Concepts

### Feature engineering

A linear model can only draw a straight line through the features it receives.
**Feature engineering** means creating new versions of variables that allow the
model to represent a different pattern.

| Feature | When it may help | What changes |
|---|---|---|
| Log | A variable has a long right tail | Large values are compressed |
| Squared term | A scatterplot bends | A straight line can become a curve |
| Category | The variable has meaningful groups or jumps | Groups are compared with a reference |
| Interaction | A slope differs between groups | Each group can have a different slope |
| Fixed effect | Unmeasured characteristics differ by place | Each place receives its own intercept |

These changes should be chosen by looking at the data first, not by trying
transformations at random.

### Logs

A log is useful when a variable is strongly right-skewed. It compresses the
long tail and treats percentage differences more similarly than equal unit
differences. For example, an increase from $20,000 to $40,000 is proportionally
much larger than an increase from $180,000 to $200,000, even though both are a
$20,000 difference.

If a predictor can equal zero, use:

```r
log1p(x)
```

This calculates $\log(x+1)$ because $\log(0)$ is negative infinity and cannot
be used in a regression.

The interpretation of a coefficient depends on which variables are logged:

| Model form | Interpretation of $\beta$ |
|---|---|
| $y \sim x$ | A one-unit increase in $x$ corresponds to $\beta$ units of change in $y$ |
| $y \sim \log(x)$ | A 1% increase in $x$ corresponds to approximately $\beta/100$ units of change in $y$ |
| $\log(y) \sim x$ | A one-unit increase in $x$ corresponds to approximately $100\beta$% change in $y$ |
| $\log(y) \sim \log(x)$ | A 1% increase in $x$ corresponds to approximately $\beta$% change in $y$ |

Logging the outcome changes the units of the residuals. An RMSE measured in
log units cannot be compared directly with an RMSE measured in dollars. The
predictions must first be transformed back to the original units with `exp()`.

### Squared terms

A squared term allows a linear model to represent a curve:

$$
\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1X + \hat{\beta}_2X^2
$$

In an R formula, the squared term is written with `I()`:

```r
y ~ x + I(x^2)
```

`I()` tells R to perform the arithmetic. The linear and squared coefficients
must be interpreted together:

- Positive linear and negative squared terms: the relationship rises and then
  levels off or turns downward.
- Negative linear and positive squared terms: the relationship falls and then
  levels off or turns upward.

The turning point is:

$$
-\frac{\beta_1}{2\beta_2}
$$

The turning point should be checked against the observed range of the data. If
it falls outside that range, the observed relationship bends but does not
actually turn. Squared terms can also make unreasonable predictions when used
far beyond the range of the training data.

### Categorical variables

Categories are useful when a variable already consists of meaningful groups or
when a numeric relationship changes in jumps instead of smoothly. Categories
can be created with `case_when()` or `cut()` and should be stored as factors.

```r
ring = case_when(
  distance < 2 ~ "Core",
  distance < 5 ~ "Inner",
  TRUE         ~ "Outer"
)
```

A factor with $k$ categories becomes $k-1$ indicator variables. The omitted
category is the **reference level**. A coefficient such as `ringOuter` is the
predicted difference between Outer and the reference group, holding the other
variables constant. It is not the predicted outcome for Outer by itself.

Changing the reference level changes the comparisons but does not change the
model's predictions. Before cross-validation, group sizes should be checked;
very small groups may be missing from a training fold.

Turning a numeric variable into categories also discards information. For
example, tracts 2.1 and 4.9 miles from City Hall become identical if both are
placed in the Inner category. Cross-validation shows whether the useful group
pattern outweighs that loss of detail.

### Interactions

An interaction is useful when the relationship between a predictor and the
outcome differs between groups. In R:

```r
y ~ x * group
```

The `*` includes `x`, `group`, and their interaction `x:group`. The coefficient
for `x` is the slope for the reference group. The slope for another group is
the reference slope plus that group's interaction coefficient.

Interactions should have a substantive reason. Crossing every variable with
every other variable adds many coefficients and can make the model unstable or
overfit the data.

### Fixed effects

A fixed effect is a categorical variable for a group of observations, often a
place such as a planning district, neighborhood, or county. It gives each place
its own intercept and absorbs characteristics that are shared within that
place but were not measured directly.

Fixed effects can capture reputation, school catchments, transit access,
history, and other location-based differences. They tell us **where** outcomes
differ, but not **why** they differ.

After adding fixed effects, another coefficient may change because it is now
estimated by comparing observations within the same place instead of across
the entire study area.

Geographic codes must be treated as labels rather than numbers. A district
code of 18 is not eighteen times district 1, so coded places should be
converted to factors.

A fixed-effects model cannot estimate an effect for a group it never saw while
training. For 10-fold cross-validation, each group should ideally have at least
about 10 observations, and more is safer. Small groups may need to be combined
with meaningful neighbors or replaced with a coarser geography.

### Using the model log

Every transformed model should be evaluated with the same rows and the same
cross-validation folds. Some changes will help, some will do almost nothing,
and others will make prediction worse. That result is the point of maintaining
a model log.

The best transformation is selected using CV RMSE rather than p-values or
in-sample $R^2$. A lower CV RMSE means the model predicts unseen observations
more accurately.

## Proving R Works

This example creates a small dataset and demonstrates logs, squared terms,
categories, and interactions without downloading external data.

```{r}
library(tidyverse)

tract_demo <- tibble(
  tract = sprintf("%03d", 1:15),
  rent = c(1050, 1120, 1180, 1260, 1350, 1430, 1510, 1580,
           1640, 1700, 1760, 1810, 1850, 1880, 1900),
  income = c(28000, 32000, 36000, 41000, 47000, 53000, 60000,
             68000, 77000, 87000, 98000, 110000, 125000, 142000, 160000),
  distance = c(0.8, 1.2, 1.7, 2.1, 2.7, 3.2, 3.8, 4.4,
               5.0, 5.6, 6.3, 7.0, 7.8, 8.7, 9.6),
  incidents = c(0, 1, 2, 1, 3, 5, 4, 7, 6, 9, 8, 12, 15, 18, 24)
) |>
  mutate(
    log_incidents = log1p(incidents),
    ring = case_when(
      distance < 2 ~ "Core",
      distance < 5 ~ "Inner",
      TRUE         ~ "Outer"
    ),
    ring = factor(ring, levels = c("Core", "Inner", "Outer"))
  )

tract_demo |>
  count(ring)
```

I can fit separate models to test different feature shapes:

```{r}
m_log <- lm(rent ~ log(income) + log_incidents, data = tract_demo)

m_squared <- lm(
  rent ~ log(income) + distance + I(distance^2),
  data = tract_demo
)

m_category <- lm(rent ~ log(income) + ring, data = tract_demo)

m_interaction <- lm(rent ~ log(income) * ring, data = tract_demo)

coef(m_log)
coef(m_squared)
coef(m_category)
coef(m_interaction)
```

The category coefficients compare Inner and Outer tracts with Core tracts, the
reference level. In the interaction model, the income slope for an Outer tract
is the `log(income)` coefficient plus the `log(income):ringOuter` coefficient.

## Reflection

I learned that feature engineering does not replace a linear model; it changes
the information and shape given to the model. Logs can make percentage changes
more meaningful, squared terms allow curves, categories represent groups, and
interactions allow slopes to vary. Fixed effects can absorb differences between
places, but they do not explain which neighborhood characteristics caused those
differences.

I also learned that a transformation that looks better in a graph may not
predict better. Each model needs to use the same rows and cross-validation
folds, and the final decision should be based on CV RMSE. Interpretation must
also match the model form: a logged variable is read in percentages, a squared
term cannot be read alone, and a categorical coefficient must name its
reference group.

## Questions I Still Have

- How large must a reduction in CV RMSE be before the added complexity is worth
  keeping?
- How should I choose category boundaries without selecting them only because
  they make the model perform better?
- When should I use a fixed effect instead of building measurable spatial
  features for the same places?
