---
title: "Week 5: Linear Regression for Prediction"
description: "Building rent models, checking residuals, and testing predictions with cross-validation."
date: 2026-09-25
---

## What We Did

We used linear regression to predict median rent across Philadelphia Census
tracts. We started with median household income as one predictor, then added
the percentage of renters, educational attainment, and the number of nearby
Indego stations.

We learned how to interpret regression coefficients, predictions, and
residuals. We also compared model fit with prediction accuracy using a
train/test split and cross-validation. Finally, we mapped the residuals to
see where rents were higher or lower than the model expected.

## Key Concepts

### Outcomes, predictors, and noise

A statistical model can be written as:

$$
Y = f(X) + \varepsilon
$$

- $Y$ is the outcome we want to explain or predict, such as median rent.
- $X$ contains the predictors, such as income or educational attainment.
- $f(X)$ is the relationship the model tries to learn.
- $\varepsilon$ represents everything affecting rent that is not captured by
  the model, including random noise and omitted characteristics.

For a simple linear regression:

$$
Y = \beta_0 + \beta_1X + \varepsilon
$$

The fitted prediction is:

$$
\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1X
$$

The intercept, $\hat{\beta}_0$, is the predicted outcome when the predictor
equals zero. The slope, $\hat{\beta}_1$, describes the predicted change in
the outcome associated with a one-unit increase in the predictor.

### Prediction and inference

Regression output can be read in two different ways:

- **Inference** focuses on $\hat{\beta}$ and asks how a predictor is associated
  with the outcome.
- **Prediction** focuses on $\hat{Y}$ and asks how accurately the model can
  predict an outcome for observations it has not seen.

A coefficient can help prediction without proving causation. For example, the
number of Indego stations may help predict rent, but the model cannot prove
that adding a station causes rent to increase. Station locations may represent
other neighborhood characteristics that the model does not include.

### Fitting a linear model in R

A simple rent model can be fitted with:

```r
m1 <- lm(med_rent ~ med_inc, data = tracts)
summary(m1)
```

The formula can be read as: predict `med_rent` using `med_inc`.

In our class example, the estimated income coefficient was approximately
`0.0081`. This represents about $0.0081 in additional predicted monthly rent
for each additional dollar of annual household income. Multiplying by $10,000
makes the result easier to understand:

$$
0.0081 \times 10{,}000 = 81
$$

A tract with $10,000 more in median household income therefore had a predicted
monthly rent approximately $81 higher.

A small p-value provides evidence that an association is statistically
detectable. It does not tell us whether the effect is large, whether it is
causal, or whether the model predicts new observations well.

### Least squares and residuals

Linear regression chooses the line that minimizes the sum of squared errors:

$$
\sum (Y_i - \hat{Y}_i)^2
$$

A **residual** is the difference between an observed value and its prediction:

$$
e_i = Y_i - \hat{Y}_i
$$

- A positive residual means the actual rent is higher than predicted.
- A negative residual means the actual rent is lower than predicted.
- A residual near zero means the prediction is close to the observed rent.

Mapping residuals can reveal spatial patterns. If nearby tracts have similar
residuals, the model may be missing a location-based characteristic.

### Rigid and flexible models

Models can be arranged from more rigid to more flexible:

| More rigid | Middle | More flexible |
|---|---|---|
| Linear regression | Lasso and ridge | Trees |
| Logistic regression | Random forest | Boosting and neural networks |

Rigid models are usually easier to explain and interpret. Flexible models can
capture more complicated relationships, but they can be harder to understand
and may overfit the training data.

A **parametric model**, such as linear regression, assumes a particular
functional form and estimates a fixed set of parameters. A **nonparametric
model** makes fewer assumptions about the functional form and can follow more
complicated patterns in the data.

### Model fit versus prediction

$R^2$ measures how much of the variation in the outcome is explained by the
model in the data used to fit it. A higher $R^2$ indicates a better in-sample
fit, but it does not guarantee better predictions for new data.

A train/test split separates the observations into two groups:

- The **training set** is used to fit the model.
- The **test set** is used to evaluate predictions on unseen observations.

In class, we used approximately 80% of the tracts for training and 20% for
testing.

Prediction error was measured using root mean squared error:

$$
RMSE = \sqrt{\frac{1}{n}\sum(Y_i-\hat{Y}_i)^2}
$$

RMSE is expressed in the same units as the outcome. For a rent model, an RMSE
of $250 means that predictions miss the observed monthly rent by approximately
$250. A lower RMSE is better.

### Cross-validation

One train/test split can give different results depending on which observations
are selected. **K-fold cross-validation** provides a more stable evaluation.

For five-fold cross-validation:

1. Divide the observations into five groups.
2. Hold out one group as the test set.
3. Train the model using the other four groups.
4. Calculate the test RMSE.
5. Repeat until each group has been used for testing.
6. Average the five RMSE values.

We used the same folds when comparing models so that the comparison was fair.

This use of **CV** means cross-validation. It is different from the
**coefficient of variation**, which measures the uncertainty of an ACS
estimate.

### Regression diagnostics

A good residual plot should look like a random cloud around zero.

**Homoscedasticity** means the residuals have a similar amount of spread across
the range of predicted values. **Heteroscedasticity** means the spread changes,
often creating a funnel shape. In that case, one RMSE may hide the fact that
the model makes much larger errors for some observations.

Noise in the outcome $Y$ becomes part of the residuals. Noise in a predictor
$X$ can flatten the estimated slope.

**Multicollinearity** occurs when predictors are strongly related to each
other. It can make individual coefficients unstable and difficult to
interpret. For a prediction task, the main question is whether the variables
improve prediction accuracy on unseen data, which can be checked with
cross-validation.

## Proving R Works

This example creates a small tract-level rent dataset and fits a linear model
that predicts median rent using median household income.

```{r}
library(tidyverse)

rent_data <- tibble(
  GEOID = c("100", "200", "300", "400", "500", "600"),
  med_rent = c(1050, 1200, 1380, 1500, 1725, 1900),
  med_inc = c(35000, 48000, 62000, 75000, 92000, 110000),
  pct_renters = c(71, 65, 59, 54, 48, 42),
  pct_ba = c(18, 25, 34, 42, 55, 68),
  n_stations = c(1, 2, 2, 3, 4, 5)
)

m1 <- lm(med_rent ~ med_inc, data = rent_data)

summary(m1)
```

I can translate the income coefficient into the predicted rent difference
associated with a $10,000 income difference:

```{r}
coef(m1)["med_inc"] * 10000
```

I can also add each tract's predicted rent and residual:

```{r}
rent_checked <- rent_data |>
  mutate(
    predicted_rent = predict(m1, newdata = rent_data),
    residual = med_rent - predicted_rent
  )

rent_checked
```

Finally, I can calculate the model's RMSE:

```{r}
sqrt(mean(rent_checked$residual^2))
```

The residuals show which observations have rents above or below the model's
expectation. The RMSE summarizes the typical size of these errors in monthly
rent dollars.

## Reflection

I learned that fitting the data and predicting new observations are related
but different goals. A model can have a higher $R^2$ because it fits the
existing observations better, but cross-validation is needed to determine
whether it also predicts unseen tracts better. I also understand that a
coefficient describes an association while a residual identifies where the
actual outcome differs from the model's prediction.

Mapping the residuals was especially useful because it showed that model
errors may form geographic clusters. If nearby tracts have rents that are
consistently higher than predicted, the model may be missing a spatial feature
such as access to Center City, universities, transit, newer housing, or other
neighborhood amenities.

## Questions I Still Have

- How do I decide whether an improvement in cross-validation RMSE is large
  enough to justify adding another predictor?
- If the residual map shows geographic clusters, what spatial variables should
  I add to the model?
- When predictors are strongly correlated, how can I tell whether their
  individual coefficients are still useful to interpret?
