---
title: "Week 4: Spatial Data, Joins, & Coordinate Systems"
description: "Connecting Census data to maps and measuring spatial relationships."
date: 2026-09-18
---

## What We Did

We joined Census income data from different years using GEOID, then learned how
spatial data stores geometry alongside a table. We discussed coordinate
reference systems (CRS), map projections, and why distance calculations need
appropriate units. We also introduced `sf` functions for finding tracts near
stations and measuring spatial relationships.

## Key Concepts

A **GEOID** identifies a Census geography. It should be stored as text so
leading zeros are preserved. When joining two tables, the GEOID columns need
to refer to the same type of geography and use matching values.

A `left_join()` keeps every row from the first table and adds matching columns
from the second. If a GEOID has no match, the new column contains `NA`. When
the first table is an `sf` object, its geometry column stays attached.

`sf` stands for **simple features**. An `sf` table contains ordinary columns
plus a geometry column with shapes such as points or polygons.

A **CRS** tells R how coordinates relate to places on Earth:

| EPSG | CRS | Units |
|---|---|---|
| 4326 | WGS 84 | Degrees |
| 4269 | NAD 83 | Degrees |
| 2272 | Pennsylvania South State Plane | US survey feet |
| 5070 | NAD 83 / Conus Albers | Meters |

A geographic CRS uses longitude and latitude. A projected CRS represents
locations on a flat map, often in feet or meters. Projections involve tradeoffs:
one may preserve local shapes, while another preserves areas or distances
better for a particular purpose.

Use `st_crs()` to check a layer's CRS. `st_set_crs()` assigns the **correct
existing CRS label** when it is missing; it does not move coordinates.
`st_transform()` converts coordinates into another CRS. I need to check the
units before making a buffer or interpreting a distance.

Some useful spatial functions are:

- `st_buffer()` creates an area around a feature.
- `st_filter()` selects features with a spatial relationship, such as tracts
  that intersect station buffers.
- `st_intersection()` creates the shared geometry where features overlap.
- `st_union()` combines geometries into one shape.
- `st_join()` adds attributes based on a spatial relationship.
- `st_distance()` calculates distances between features.

## Proving R Works

This example joins income data from two years by GEOID. The 2023 table goes
first, so all three of its GEOIDs remain. GEOID `"300"` has no 2015 match and
receives `NA`.

```{r}
library(dplyr)
library(tibble)

income_23 <- tibble(
  GEOID = c("100", "200", "300"),
  med_inc_23 = c(82000, 31500, 45000)
)

income_15 <- tibble(
  GEOID = c("100", "200", "400"),
  med_inc_15 = c(61000, 27000, 39000)
)

income_joined <- left_join(income_23, income_15, by = "GEOID")

income_joined
```

I can check the join by counting rows and missing values:

```{r}
income_joined |>
  summarize(
    rows = n(),
    missing_2015_income = sum(is.na(med_inc_15))
  )
```

The result should show **3 rows** and **1 missing 2015 income value**.
This example uses ordinary tables; a spatial version would also have a
geometry column.

## Reflection

I learned that joining Census data to a map depends on both the table key and
the spatial data. A join can run successfully while still leaving unmatched
GEOIDs, so I should check the row count and missing values afterward. I also
understand why I need to check the CRS before making a half-mile buffer:
the number supplied to `st_buffer()` only makes sense when I know the layer's
distance units.

## Questions I Still Have

- If Census tract boundaries change between 2015 and 2023, is matching by
  GEOID alone enough to compare income across years?
- Should a tract count as near a station when any part of it touches a
  half-mile buffer, or only when its centroid falls inside?
- When `st_distance()` returns distances to many stations, how do I find
  just the nearest station for each tract?
