---
title: "Assignment 3: Verify the Claim: Poverty Change in Philadelphia"
author: "Terry Xu"
date: 2026-09-24
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
---
## Article

[New Census figures on Philly neighborhoods show inequality, high numbers of whites living in poverty](https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/)
```{r}
#| label: setup
#| include: false

library(tidyverse)
library(tidycensus)
library(gt)
```
## Part 1: Reconstructing the Article's Data

The article compares Philadelphia neighborhood poverty using the 2006–2010 and
2013–2017 ACS 5-year estimates. I use `survey = "acs5"` with `year = 2010` and
`year = 2017` to retrieve those periods. The periods do not overlap, so they
do not share collection years. For the change test, I use the standard
approximation that the estimates from the two periods are independent.

I use table B17001: `B17001_002` counts people below the poverty level, and
`B17001_001` counts people for whom poverty status is determined. Dividing
these gives the poverty rate. I assume this measure corresponds to the
article's overall poverty measure. Because its neighborhood-to-tract definitions
are unavailable, I compare Philadelphia census tracts rather than reconstructing
its neighborhood totals. I match tracts by GEOID and assume matched tracts
represent comparable areas; matching identifiers alone does not verify unchanged
boundaries.
```{r}
#| label: pull-poverty-data
#| message: false
#| warning: false

poverty_10 <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(
    poor = "B17001_002",
    poverty_universe = "B17001_001"
  ),
  year = 2010,
  survey = "acs5",
  output = "wide"
)

poverty_17 <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(
    poor = "B17001_002",
    poverty_universe = "B17001_001"
  ),
  year = 2017,
  survey = "acs5",
  output = "wide"
)

dim(poverty_10)
dim(poverty_17)
```
```{r}
#| label: join-periods

poverty_10 <- poverty_10 %>%
  select(
    GEOID,
    poor_10E = poorE,
    poor_10M = poorM,
    universe_10E = poverty_universeE,
    universe_10M = poverty_universeM
  )

poverty_17 <- poverty_17 %>%
  select(
    GEOID,
    NAME,
    poor_17E = poorE,
    poor_17M = poorM,
    universe_17E = poverty_universeE,
    universe_17M = poverty_universeM
  )

poverty <- poverty_17 %>%
  left_join(poverty_10, by = "GEOID")

poverty %>%
  summarize(
    n_tracts = n(),
    no_2010_match = sum(is.na(poor_10E))
  )
```

## Part 2: Testing Poverty Change
```{r}
#| label: calculate-poverty-rates

poverty <- poverty %>%
  mutate(
    poverty_rate_10 = poor_10E / universe_10E,
    poverty_moe_10 = moe_prop(
      poor_10E,
      universe_10E,
      poor_10M,
      universe_10M
    ),
    poverty_se_10 = poverty_moe_10 / 1.645,
    poverty_cv_10 = poverty_se_10 / poverty_rate_10 * 100,

    poverty_rate_17 = poor_17E / universe_17E,
    poverty_moe_17 = moe_prop(
      poor_17E,
      universe_17E,
      poor_17M,
      universe_17M
    ),
    poverty_se_17 = poverty_moe_17 / 1.645,
    poverty_cv_17 = poverty_se_17 / poverty_rate_17 * 100,

    reliability_17 = case_when(
      is.na(poverty_cv_17) ~ NA_character_,
      poverty_cv_17 < 12  ~ "Reliable",
      poverty_cv_17 <= 40 ~ "Somewhat reliable",
      TRUE                ~ "Unreliable"
    )
  )

poverty %>%
  summarize(
    n_tracts = n(),
    missing_2010_rate = sum(is.na(poverty_rate_10)),
    missing_2017_rate = sum(is.na(poverty_rate_17))
  )
```
```{r}
#| label: inspect-missing-rates

poverty %>%
  filter(
    is.na(poverty_rate_10) | is.na(poverty_rate_17)
  ) %>%
  select(
    GEOID,
    NAME,
    poor_10E,
    universe_10E,
    poor_17E,
    universe_17E
  ) %>%
  print(n = Inf, width = Inf)
```
```{r}
#| label: keep-comparable-tracts

poverty_compare <- poverty %>%
  filter(
    universe_10E > 0,
    universe_17E > 0
  )

nrow(poverty_compare)
```
```{r}
#| label: test-poverty-change

poverty_compare <- poverty_compare %>%
  mutate(
    change_pp = (poverty_rate_17 - poverty_rate_10) * 100,
    change_moe_pp = sqrt(
      poverty_moe_10^2 + poverty_moe_17^2
    ) * 100,
    lower = change_pp - change_moe_pp,
    upper = change_pp + change_moe_pp,
    change_result = case_when(
      is.na(lower) | is.na(upper) ~ "Cannot test",
      lower > 0 ~ "Detectable increase",
      upper < 0 ~ "Detectable decrease",
      TRUE ~ "No detectable change"
    )
  )

poverty_compare %>%
  count(change_result)
```
```{r}
#| label: reliability-both-periods

poverty_compare <- poverty_compare %>%
  mutate(
    reliability_10 = case_when(
      is.na(poverty_cv_10) ~ "Cannot assess",
      poverty_cv_10 < 12 ~ "Reliable",
      poverty_cv_10 <= 40 ~ "Somewhat reliable",
      TRUE ~ "Unreliable"
    ),
    reliability_17 = case_when(
      is.na(poverty_cv_17) ~ "Cannot assess",
      poverty_cv_17 < 12 ~ "Reliable",
      poverty_cv_17 <= 40 ~ "Somewhat reliable",
      TRUE ~ "Unreliable"
    )
  )

poverty_compare %>%
  count(reliability_10)

poverty_compare %>%
  count(reliability_17)
```
```{r}
#| label: reliability-table

reliability_counts_10 <- poverty_compare %>%
  count(reliability_10) %>%
  rename(
    reliability = reliability_10,
    tracts_2010 = n
  )

reliability_counts_17 <- poverty_compare %>%
  count(reliability_17) %>%
  rename(
    reliability = reliability_17,
    tracts_2017 = n
  )

reliability_table <- reliability_counts_10 %>%
  left_join(reliability_counts_17, by = "reliability") %>%
  mutate(
    symbol = case_when(
      reliability == "Reliable" ~ "✓",
      reliability == "Somewhat reliable" ~ "~",
      reliability == "Unreliable" ~ "!"
    )
  ) %>%
  select(symbol, reliability, tracts_2010, tracts_2017)

reliability_table %>%
  gt() %>%
  tab_header(
    title = "Reliability of Philadelphia Tract Poverty Rates",
    subtitle = "376 tracts included in the change comparison"
  ) %>%
  cols_label(
    symbol = "Symbol",
    reliability = "CV category",
    tracts_2010 = "2006–2010",
    tracts_2017 = "2013–2017"
  ) %>%
  tab_source_note(
    source_note = "Source: ACS 5-year estimates, table B17001. Jurjevich et al. (2018) thresholds: ✓ Reliable, CV < 12%; ~ Somewhat reliable, CV 12–40%; ! Unreliable, CV > 40%."
  )
```
```{r}
#| label: change-error-bars
#| fig-width: 9
#| fig-height: 8

largest_increases <- poverty_compare %>%
  arrange(desc(change_pp)) %>%
  head(10)

largest_decreases <- poverty_compare %>%
  arrange(change_pp) %>%
  head(10)

plot_tracts <- bind_rows(largest_increases, largest_decreases)

ggplot(
  plot_tracts,
  aes(x = reorder(GEOID, change_pp), y = change_pp)
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.2
  ) +
  geom_point(aes(color = change_result), size = 2.5) +
  coord_flip() +
  labs(
    title = "Largest Estimated Changes in Philadelphia Tract Poverty",
    subtitle = "2006–2010 to 2013–2017: 10 largest increases and 10 largest decreases",
    x = "Census tract GEOID",
    y = "Change in poverty rate (percentage points)",
    color = "90% confidence test",
    caption = "Source: ACS 5-year estimates, B17001. Bars show 90% confidence intervals for change."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
This figure shows the ten largest estimated increases and ten largest estimated
decreases, rather than all 376 tracts. Error bars that cross zero indicate changes
that are not statistically detectable at 90% confidence.

Tract 9891 illustrates the importance of uncertainty. Its estimated poverty rate
increased by 23.3 percentage points, but the change's margin of error is about
221 percentage points. Its poverty-status population estimates were only 68 and
53, with margins of error larger than those estimates. The resulting approximate
interval extends beyond possible changes, indicating that this comparison is too
imprecise to support a conclusion about the direction of change.

```{r}
#| label: change-summary-chart
#| fig-width: 8
#| fig-height: 5

change_counts <- poverty_compare %>%
  count(change_result)

ggplot(
  change_counts,
  aes(x = change_result, y = n)
) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(
    title = "Most Philadelphia Tracts Had No Detectable Poverty-Rate Change",
    subtitle = "2006–2010 compared with 2013–2017, using 90% confidence",
    x = NULL,
    y = "Number of census tracts",
    caption = "Source: ACS 5-year estimates, B17001. Includes 376 tracts with defined rates in both periods."
  ) +
  theme_minimal()
```
### What the results show

Of the 376 tracts included in the comparison, 36 showed a statistically detectable
increase in poverty and 27 showed a detectable decrease at 90% confidence.
The remaining 313 tracts, or 83.2%, showed no statistically detectable change.
This does not mean their poverty rates stayed exactly the same. It means the
estimated changes were not large enough relative to their uncertainty to
distinguish them from zero.

The reliability table also shows that most tract poverty rates were only
somewhat reliable. In 2006–2010, 11 tracts were classified as reliable, 321 as
somewhat reliable, and 44 as unreliable. In 2013–2017, those counts were 22,
332, and 22. These categories describe the precision of each period's poverty
rate; they do not directly determine whether the change between periods is
statistically detectable.

Overall, the results show increases in some tracts and decreases in others.
They do not establish what happened in the specific neighborhoods named in the
article, because the article's neighborhood-to-tract definitions are unavailable.
The tract counts also do not measure the change in Philadelphia's overall poverty
rate, since tracts contain different numbers of people.

## Part 3: Which Claims Does the Evidence Support?

**Supported:** The article reports that poverty increased in some parts of
Philadelphia. My tract-level analysis supports that broad pattern: 36 tracts
showed statistically detectable increases between 2006–2010 and 2013–2017.
However, 27 tracts showed detectable decreases, so the direction of change
varied across the city. **Not supported by this analysis:** Treating every
estimated increase as clear evidence that conditions worsened would overstate
the findings. For example, Tract 9891 had an estimated increase of 23.3 percentage
points, but its extremely large margin of error prevents a confident conclusion
about the direction of change. Across all 376 comparable tracts, 313 changes
were not statistically distinguishable from zero. This does not prove that
poverty was unchanged in those places.

**Cannot be determined:** I cannot verify the article's claim that Eastwick's
poverty rate rose from 12% to 26%, or that it had the largest neighborhood
increase, because the article does not identify the tracts used to define each
neighborhood. The same limitation prevents verification of Fairhill's reported
61% poverty rate. My analysis also does not test the article's claim about
increasing poverty among white residents: the variables used here describe
poverty across the full poverty-status population, not a specific racial group.
Finally, these comparisons cannot establish the suggested causes of change,
such as opioid use or migration. Those explanations would require additional
evidence.

## Part 4: A More Defensible Account

Philadelphia's poverty trends varied across census tracts between the
2006–2010 and 2013–2017 American Community Survey periods. Among 376 tracts
with usable poverty rates in both periods, 36 showed increases and 27 showed
decreases that were statistically detectable at 90% confidence. For the other
313, uncertainty was too large to establish a change. These results indicate
worsening poverty in some areas and improvement in others, rather than a
uniform trend. They also show why a large estimated increase should not
automatically be treated as a clear deterioration. The article's exact figures
for neighborhoods such as Eastwick and Fairhill cannot be verified through this
tract-level analysis without knowing how the neighborhood boundaries were
defined.






