---
title: "Assignment 4: Building Features, Watching What Happens"
subtitle: "MUSA 5080 · Integrates Weeks 1–5 · Due Friday, October 16"
author: "Terry Boxuan Xu"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
execute:
  warning: false
  message: false
freeze: false
---

## What this assignment is (and isn't)

This is **not a competition.** There's no leaderboard and no target RMSE. The goal is to practice the full prediction workflow on a question you choose, and to **find out what actually helps.** Some features you're sure will help won't, and some you add on a whim will. Both are findings.

You'll:

1. Choose a tract-level outcome for Philadelphia and check whether it's reliable enough to model (Weeks 2–3).
2. Fit a baseline model and cross-validate it (Week 5).
3. Pull data from Philadelphia's open data portal through its API and build spatial features: counts, densities, and distances (Week 4, plus one new tool).
4. Transform variables with logs, squared terms, and categories. **Watch the October 9 recorded lecture before Part 5.**
5. Keep a **model log** the whole way through, and compare where your first and final models go wrong.

**Timing.** Parts 1–4 use only what we've covered through Week 5. You can finish them before the October 9 lecture. Part 5 is new material; the lecture explains it, and the instructions below give you the essentials.

**Type the code yourself.** The worked examples show the pattern. Your job is to apply it to your outcome and your data choices.

### Three rules that apply everywhere

::: {.callout-important}
## 1. Don't drop rows
Every model is cross-validated on the same tracts with the same folds, so the rows in your model log are comparable. If a new variable has missing values, fill them (for example, 0 for a count, or the median with a note) or choose a different variable. The `cv_rmse()` helper below will stop and tell you if you break this rule.

## 2. Stay inside the time window
Your outcome comes from the 2019–2023 ACS. Pull open data from **2023 or earlier** (2021–2023 is ideal). Using 2025 data to predict a 2023 outcome uses information from the future, which a real prediction model would never have.

## 3. No leakage
Don't use any variable that's built from your outcome or measures the same thing another way. Ask yourself: *would an agency trying to predict this outcome for a tract have this variable?* If the answer is "only because it's basically the outcome," leave it out.
:::

---

## Part 0: Setup

```{r}
#| label: setup
library(tidyverse)
library(tidycensus)
library(sf)
library(caret)

Sys.getenv("CENSUS_API_KEY") != ""   # should print TRUE
```

### The cross-validation helper

This function runs 10-fold cross-validation with the same seed every time, so every model in your log is tested on the same folds. **You don't need to change it.** Read it once so you know what it does.

```{r}
#| label: cv-helper
ctrl <- trainControl(method = "cv", number = 10)

cv_rmse <- function(formula, data) {
  df <- if (inherits(data, "sf")) st_drop_geometry(data) else data
  vars <- all.vars(formula)
  bad <- vars[sapply(vars, function(v) any(is.na(df[[v]])))]
  if (length(bad) > 0) {
    stop("These variables have NAs: ", paste(bad, collapse = ", "),
         ". Fill them or choose another variable (Rule 1: don't drop rows).")
  }
  set.seed(5080)
  fit <- train(formula, data = df, method = "lm", trControl = ctrl)
  fit$results$RMSE
}
```

### The model log

Every time you fit a new model, add a row. By the end, this table is the story of your assignment.

```{r}
#| label: log-start
model_log <- tibble(
  model   = character(),
  added   = character(),
  cv_rmse = numeric(),
  notes   = character()
)
```

---

## Part 1: Choose your outcome

Pick a **tract-level ACS variable for Philadelphia** that matters for policy. Two outcomes are off limits: median gross rent (we did it in lab) and median home value (save it for A4).

This analysis predicts the percentage of households without access to a vehicle in each Philadelphia Census tract. This outcome is relevant to transportation planning because neighborhoods with more households without vehicles may depend more heavily on public transit, walking, and biking. The outcome uses ACS table B08201.

Some ideas if you're stuck:

| Outcome | Why an agency might care | ACS table |
|---|---|---|
| % of households with no vehicle | Transit planning | B08201 or B25044 |
| Mean travel time to work | Access to jobs | B08013_001 ÷ B08303_001 |
| % of people below poverty | Targeting services | B17001 |
| % of housing units vacant | Housing and blight | B25002 |
| % of households overcrowded | Housing conditions | B25014 |

Use `load_variables(2023, "acs5")` to browse and find exact variable codes.

### 1A: Pull your outcome and your baseline predictors

Choose **three to five ACS predictors** for your baseline model. They should be things that plausibly relate to your outcome, but **none of them may be built from your outcome** (Rule 3).

```{r}
#| label: pull
philly_tracts <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(
    total_pop = "B01003_001",
    hh_total = "B08201_001",
    hh_no_vehicle = "B08201_002",
    med_inc = "B19013_001",
    occupied = "B25003_001",
    renter = "B25003_003",
    edu_total = "B15003_001",
    bachelors = "B15003_022",
    masters = "B15003_023",
    professional = "B15003_024",
    doctorate = "B15003_025"
  ),
  year = 2023,
  output = "wide",
  geometry = TRUE
) %>%
  st_transform(2272)
```

```{r}
#| label: build-outcome
philly_tracts <- philly_tracts %>%
  mutate(
    # build your outcome, e.g. a percentage:
    y = hh_no_vehicleE / hh_totalE * 100,
    # and its reliability. For a percentage, use tidycensus::moe_prop();
    # for a median or count, use the E and M columns directly.
    y_moe = tidycensus::moe_prop(
      hh_no_vehicleE,
      hh_totalE,
      hh_no_vehicleM,
      hh_totalM
    ) * 100,
    y_cv  = (y_moe / 1.645) / y * 100,
    # your predictors:
    pct_renter = renterE / occupiedE * 100,
    pct_ba = (
      bachelorsE + mastersE + professionalE + doctorateE
    ) / edu_totalE * 100
  )
```

### 1B: Is it reliable enough to model?

Before you commit, check three things.

```{r}
#| label: reliability
# How many tracts have no value for your outcome?
sum(is.na(philly_tracts$y))

# How reliable is it?
summary(philly_tracts$y_cv)

philly_tracts %>%
  mutate(reliability = case_when(
    y_cv < 12  ~ "High",
    y_cv <= 40 ~ "Moderate",
    TRUE       ~ "Low"
  )) %>%
  st_drop_geometry() %>%
  count(reliability)

# What does the distribution look like?
ggplot(philly_tracts, aes(x = y)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

If more than about a third of tracts have low reliability, consider a different outcome, or one built from larger counts. Write your reasoning either way.

**Answer (2–3 sentences):** What is your outcome, why would an agency want to predict it, and how reliable is it?

The outcome is the percentage of households without access to a vehicle in each Philadelphia Census tract. A transportation agency may want to predict this percentage to identify communities that depend more heavily on public transit, walking, or biking. Most tracts have moderately reliable estimates, with 29 classified as High reliability, 313 as Moderate, and 66 as Low, so I decided to continue using this outcome while recognizing that some tract estimates are uncertain.

### 1C: Decide which tracts to model

Filter once, here, and never again. Every later step uses `tracts`.

```{r}
#| label: filter
tracts <- philly_tracts %>%
  filter(
    total_popE > 0,
    !is.na(y),
    !is.na(med_incE),
    !is.na(pct_renter),
    !is.na(pct_ba)
  )

nrow(philly_tracts)
nrow(tracts)
```

**Answer:** How many tracts did you lose, and what kinds of places are they?

I removed 27 tracts, leaving 381 tracts for every model. Most of the excluded areas were zero-population 9800-series tracts, while several populated or very small-population tracts were removed because the ACS did not provide a usable outcome or baseline predictor, especially median household income.

---

## Part 2: Baseline model

### 2A: Look first

Plot your outcome against each predictor before modeling (the Week 3 habit).

```{r}
#| label: look

ggplot(tracts, aes(x = med_incE, y = y)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) + 
  theme_minimal()

ggplot(tracts, aes(x = pct_renter, y = y)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) + 
  theme_minimal()

ggplot(tracts, aes(x = pct_ba, y = y)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) + 
  theme_minimal()
```

### 2B: Fit and log

```{r}
#| label: m1
f1 <- y ~ med_incE + pct_renter + pct_ba

m1 <- lm(f1, data = tracts)
summary(m1)

model_log <- model_log %>%
  add_row(
    model = "m1", 
    added = "ACS baseline",
    cv_rmse = cv_rmse(f1, tracts),
    notes = "Income, renter share, and education"
  )

model_log
```

On tracts it hasn't seen, the baseline model misses by about 12.5 percentage points.

### 2C: Where is the baseline wrong?

Save this map. You'll compare it to your final model's map in Part 6.

```{r}
#| label: resid-map-1
tracts <- tracts %>% mutate(resid_m1 = residuals(m1))

ggplot(tracts) +
  geom_sf(aes(fill = resid_m1), color = NA) +
  scale_fill_gradient2(low = "#2166ac", mid = "grey95", high = "#b2182b", midpoint = 0) +
  theme_void() +
  labs(title = "Baseline model: where it's wrong",
       subtitle = "Red = actual higher than predicted",
       fill = "Residual")
```

**Answer:** Where do the residuals cluster? What might those places have in common that your model doesn't know about? **Use this answer to choose your open data layers in Part 3.**

The residuals are spatially clustered rather than randomly scattered. Downtown has several red tracts, possibly because its density, transit access, and limited parking make living without a vehicle easier than the baseline model predicts. Red clusters in West and North Philadelphia may also be influenced by the large student populations around Penn and Temple, which the model does not directly measure.

---

## Part 3: Getting data from Philadelphia's open data portal

Many of the city's datasets on [OpenDataPhilly](https://opendataphilly.org) are served through a **Carto SQL API**. You send a SQL query in a URL, and it sends back data. Ask for `format=GeoJSON` and `st_read()` reads the result directly, just like the Indego feed in Week 4.

The pattern:

```
https://phl.carto.com/api/v2/sql?format=GeoJSON&q=SELECT ... FROM table_name WHERE ...
```

Three pieces of SQL are all you need:

- `SELECT the_geom, column1, column2` chooses which columns come back. **Always include `the_geom`**, or you get no geometry.
- `FROM table_name` chooses the dataset.
- `WHERE ...` filters rows. **Filter in the query, not in R.** Some tables have millions of rows, and pulling everything will be slow or fail.

This helper builds the URL for you. Spaces and symbols in a URL need encoding, and `URLencode()` handles that.

```{r}
#| label: carto-helper
carto_get <- function(sql) {
  url <- paste0("https://phl.carto.com/api/v2/sql?format=GeoJSON&q=", URLencode(sql))
  st_read(url, quiet = TRUE)
}
```

### Finding a table name

On a dataset's OpenDataPhilly page, look for the API link. The table name is the word after `FROM` in the example query. To see what values a column takes before you filter on it, ask for a count in CSV format:

```{r}
#| label: explore-values
# What kinds of crime are in the incidents table in 2023?
url <- paste0("https://phl.carto.com/api/v2/sql?format=csv&q=",
              URLencode("SELECT text_general_code, COUNT(*) AS n
                         FROM incidents_part1_part2
                         WHERE dispatch_date >= '2023-01-01' AND dispatch_date < '2024-01-01'
                         GROUP BY text_general_code ORDER BY n DESC"))
read_csv(url)
```

### Worked example: residential burglaries in 2023

```{r}
#| label: worked-burglary
burglary <- carto_get(
  "SELECT the_geom, text_general_code, dispatch_date
   FROM incidents_part1_part2
   WHERE dispatch_date >= '2023-01-01' AND dispatch_date < '2024-01-01'
     AND text_general_code = 'Burglary Residential'"
) %>%
  st_transform(st_crs(tracts))    # Carto returns EPSG 4326; translate immediately

nrow(burglary)
```

::: {.callout-warning}
## Check for missing geometry
Some records have no location. Drop them with `filter(!st_is_empty(burglary))`, and count how many you lost. Every point you drop is a point that won't be counted.
:::

### Other layers to try

These tables were available on Carto as of this fall. If a query fails, check the dataset's OpenDataPhilly page for the current table name.

| Data | Table | Filter on | Notes |
|---|---|---|---|
| Crime incidents | `incidents_part1_part2` | `dispatch_date`, `text_general_code` | Pick specific offense types |
| Shooting victims | `shootings` | `date_` | Small enough to pull several years |
| 311 requests | `public_cases_fc` | `requested_datetime`, `service_name` | Use the count query above to see service names |
| Indego stations | *(not Carto)* `https://www.rideindego.com/stations/json/` | — | The Week 4 feed |

Datasets not on Carto often have a GeoJSON download link on their OpenDataPhilly page. `st_read()` reads those URLs too.
For example, navigage to https://opendataphilly.org/, find a dataset that you'd like to import. I found city landmarks. On the page there is a link to a GeoJSON file. I simply right-clicked on that hyperlink and copied it and then used st_read() and pasted the link: 

landmarks <- st_read("https://hub.arcgis.com/api/v3/datasets/6b2fe4a0fbe14e0a8e6afceaea230b8e_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1")

### Your turn

Pull **at least two layers** that you think relate to your outcome. Base your choice on your Part 2C answer.

```{r}
#| label: my-layers
layer1 <- carto_get(
  "SELECT the_geom, text_general_code, dispatch_date
   FROM incidents_part1_part2
   WHERE dispatch_date >= '2023-01-01'
     AND dispatch_date < '2024-01-01'
     AND text_general_code = 'Burglary Residential'"
) %>%
  st_transform(st_crs(tracts))

layer2 <- st_read(
  "https://www.rideindego.com/stations/json/",
  quiet = TRUE
) %>%
  st_transform(st_crs(tracts))
```

```{r}
#| label: layer-check

sum(st_is_empty(layer1))
sum(st_is_empty(layer2))

layer1 <- layer1 %>%
  filter(!st_is_empty(layer1))

layer2 <- layer2 %>%
  filter(!st_is_empty(layer2))

nrow(layer1)
nrow(layer2)
```

**Answer:** Why these layers? What do you expect them to capture that the ACS variables don't?

I chose residential burglaries and Indego stations because they represent neighborhood conditions that the baseline ACS variables do not directly measure. Residential burglaries provide a measure of public-safety conditions, while Indego stations represent access to bike sharing and an alternative to private vehicles. I expect Indego access to help explain some of the higher-than-predicted percentages of households without vehicles in central and university-area tracts.

---

## Part 4: Building spatial features

Every feature below reuses a Week 4 tool. The worked examples use burglaries and Indego stations; you build the same types from your own layers.

### 4A: Counting

How many burglaries fall in each tract?

```{r}
#| label: count
tracts <- tracts %>%
  mutate(n_burglary = lengths(st_intersects(., burglary)))

summary(tracts$n_burglary)
```

A tract with no burglaries gets 0, not NA, so Rule 1 is safe.

### 4B: Density: don't skip this

Raw counts mostly measure **tract size**. A big tract has more of everything. Normalize by area, by population, or by housing units, depending on what makes sense for the thing you're counting.

```{r}
#| label: density
tracts <- tracts %>%
  mutate(
    area_sqmi         = as.numeric(st_area(.)) / 27878400,   # sq ft in a sq mile
    burglary_per_sqmi = n_burglary / area_sqmi,
    burglary_per_1k   = n_burglary / total_popE * 1000
  )
```

::: {.callout-note}
## Per capita in low-population tracts
If a tract has 40 residents and 3 burglaries, its rate is 75 per 1,000, which is a huge number built on tiny counts. Look at the tracts with the highest per-capita rates before you trust them. This is the Week 3 lesson about small denominators again.
:::

### 4C: Measuring distance to the nearest thing

How far is each tract from its nearest Indego station? Use the Week 4 pattern: find the nearest feature, then measure to it **element by element**.

```{r}
#| label: nearest
stations <- st_read("https://www.rideindego.com/stations/json/", quiet = TRUE) %>%
  st_transform(st_crs(tracts))

tract_pts <- st_point_on_surface(tracts)   # always inside the tract, unlike a centroid

nearest_idx <- st_nearest_feature(tract_pts, stations)

tracts <- tracts %>%
  mutate(dist_station_mi = as.numeric(
    st_distance(tract_pts, stations[nearest_idx, ], by_element = TRUE)) / 5280)
```

### 4D: Measuring distance to one place

Sometimes the "thing" is a single location, such as City Hall as a stand-in for the center of the regional economy.

```{r}
#| label: city-hall
city_hall <- st_sfc(st_point(c(-75.1636, 39.9526)), crs = 4326) %>%
  st_transform(st_crs(tracts))

tracts <- tracts %>%
  mutate(dist_ch_mi = as.numeric(st_distance(tract_pts, city_hall)) / 5280)
```

### 4E: Your turn

Build **at least three spatial features** from your own layers, and use at least two types (count, density, nearest distance, distance to a place). After **each one**, fit a model, cross-validate it, and log it.

```{r}
#| label: my-spatial
tracts <- tracts %>%
  mutate(
    n_burglary = lengths(st_intersects(., layer1))
  )

f2 <- y ~ med_incE + pct_renter + pct_ba + n_burglary

model_log <- model_log %>%
  add_row(model = "m2", added = "Burglary count",
          cv_rmse = cv_rmse(f2, tracts),
          notes = "Raw count of 2023 residential burglaries"
        )

# repeat for each feature
model_log
```

```{r}
tracts <- tracts %>%
  mutate(
    burglary_per_sqmi = n_burglary / area_sqmi
  )

f3 <- y ~ med_incE + pct_renter + pct_ba + burglary_per_sqmi

model_log <- model_log %>%
  add_row(
    model = "m3",
    added = "Burglary density",
    cv_rmse = cv_rmse(f3, tracts),
    notes = "Residential burglaries per square mile"
  )

model_log
```

```{r}
f4 <- y ~ med_incE + pct_renter + pct_ba + dist_station_mi

model_log <- model_log %>%
  add_row(
    model = "m4",
    added = "Distance to Indego",
    cv_rmse = cv_rmse(f4, tracts),
    notes = "Distance to nearest Indego station in miles"
  )

model_log
```

**Answer:** Which spatial feature helped most? Did any feature you expected to help do nothing, or make CV RMSE worse? Why might that be?

Distance to the nearest Indego station helped the most, reducing CV RMSE from 12.46 in the baseline model to 11.22. Burglary density also improved prediction and performed better than the raw burglary count, probably because density accounts for differences in tract size. None of the three features made prediction worse, but the raw count helped the least because larger tracts can contain more incidents simply because they cover more area.

---

## Part 5: Transforming features

**Watch the October 9 recorded lecture before you start this part.** Below is a summary to work from.

So far, every predictor has entered the model as a straight line: each additional unit of X changes predicted Y by the same amount, everywhere. Transformations let the model bend. For prediction, the question is simply **does the bend lower CV RMSE?** For interpretation, each transformation changes what a coefficient means, so you need to know how to read it.

::: {.callout-tip}
## Transform predictors, not your outcome
Keep your outcome `y` untransformed in this assignment. If you model `log(y)`, your CV RMSE comes out in log units and can't be compared to the other rows in your log. (Predicting a logged outcome well requires back-transforming, which is an A4 topic.)
:::

### 5A: Logs, for skewed predictors

Many urban variables are skewed: a few tracts have huge values, and most have small ones. Income, density, and counts are typical. A log compresses the long tail, so a model can use the difference between 1 and 10 as well as the difference between 100 and 1,000.

```{r}
#| label: log-look
ggplot(tracts, aes(x = burglary_per_sqmi)) + geom_histogram(bins = 30) + theme_minimal()
ggplot(tracts, aes(x = log1p(burglary_per_sqmi))) + geom_histogram(bins = 30) + theme_minimal()
```

::: {.callout-warning}
## log(0) doesn't exist
`log(0)` is `-Inf`, and it will break your model. Counts and densities often include zeros. Use `log1p(x)`, which computes `log(x + 1)`, for anything that can be 0.
:::

**Reading the coefficient:** when a predictor is logged, a **1% increase** in X goes with a change of about **β/100** in Y, in Y's units. With `log1p`, this reading is approximate for small values of X.

### 5B: Squared terms, for curves

If your scatterplot bends (rising and then flattening, or falling and then rising), a squared term lets the line curve. Write it with `I()`, which tells R to do the arithmetic rather than read `^` as formula syntax.

```{r}
#| label: squared-example
m_sq <- lm(y ~ dist_ch_mi + I(dist_ch_mi^2), data = tracts)
summary(m_sq)
```

**Reading the coefficient:** you can't read either coefficient alone anymore. The effect of X depends on where you are along X. A positive linear term with a negative squared term means Y rises with X and then levels off or turns down. Plot the fitted curve to see it:

```{r}
#| label: squared-plot
ggplot(tracts, aes(x = dist_ch_mi, y = y)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), se = FALSE, color = "red") +
  theme_minimal()
```

### 5C: Categorical variables

Sometimes a number works better as groups, either because the relationship isn't smooth or because the categories are meaningful. You can make categories from any variable with `case_when()` or `cut()`, and then make sure R treats them as a **factor**.

```{r}
#| label: categorical-example
tracts <- tracts %>%
  mutate(
    ring = case_when(
      dist_ch_mi < 2 ~ "Core",
      dist_ch_mi < 5 ~ "Inner",
      TRUE           ~ "Outer"
    ),
    ring = factor(ring, levels = c("Core", "Inner", "Outer"))
  )

tracts %>% st_drop_geometry() %>% count(ring)

m_cat <- lm(y ~ ring, data = tracts)
summary(m_cat)
```

**Reading the coefficients:** R turns a factor with k categories into k − 1 dummy variables. The category left out is the **reference level** (here, "Core," the first level). Each coefficient is the **difference from the reference**: `ringOuter` is how much higher or lower predicted Y is in outer tracts *compared to core tracts*, holding everything else in the model constant. To change the reference, use `relevel(ring, ref = "Outer")`.

::: {.callout-note}
## Check your group sizes
If a category has only a handful of tracts, its coefficient is unstable, and in cross-validation some folds may not contain it at all. Aim for at least 20 tracts per category.
:::

### 5D: Your turn

Try **at least one of each**: a log, a squared term, and a categorical variable. Choose which variables to transform based on your plots, not at random. Log each model.

```{r}
#| label: my-transforms
f_log <- y ~ med_incE + pct_renter + pct_ba +
  log1p(burglary_per_sqmi)

model_log <- model_log %>%
  add_row(model = "m5", added = "Log burglary density",
          cv_rmse = cv_rmse(f_log, tracts),
          notes = "Logged residential burglaries per square mile")

# squared term

# categorical

model_log
```

```{r}
f_sq <- y ~ med_incE + pct_renter + pct_ba +
  dist_ch_mi + I(dist_ch_mi^2)

model_log <- model_log %>%
  add_row(
    model = "m6",
    added = "Squared City Hall distance",
    cv_rmse = cv_rmse(f_sq, tracts),
    notes = "Distance to City Hall with a curved relationship"
  )

model_log
```

```{r}
f_cat <- y ~ med_incE + pct_renter + pct_ba + ring

m7 <- lm(f_cat, data = tracts)
summary(m7)

model_log <- model_log %>%
  add_row(
    model = "m7",
    added = "City Hall distance categories",
    cv_rmse = cv_rmse(f_cat, tracts),
    notes = "Core, Inner, and Outer distance rings"
  )

model_log
```

**Answer:** Which transformation helped most? For your categorical variable, write one sentence reading one of its coefficients correctly, naming the reference level.

The squared City Hall distance transformation helped the most, producing the lowest CV RMSE of 9.76, compared with 12.46 for the baseline model. Using Core tracts as the reference level, Outer tracts are predicted to have about 21.95 percentage points fewer households without vehicles, holding median income, renter percentage, and educational attainment constant.

---

## Part 6: Your final model

Combine the features and transformations that earned their place, **judged by CV RMSE, not p-values.**

```{r}
#| label: final
f_final <- y ~ med_incE + pct_renter + pct_ba +
  burglary_per_sqmi +
  dist_station_mi +
  dist_ch_mi + I(dist_ch_mi^2)

m_final <- lm(f_final, data = tracts)
summary(m_final)

model_log <- model_log %>%
  add_row(model = "final", added = "Best spatial features and squared distance",
          cv_rmse = cv_rmse(f_final, tracts),
          notes = "Raw burglary density, Indego distance, and curved City Hall distance")

model_log
```

### Where is the final model wrong?

```{r}
#| label: resid-map-2
tracts <- tracts %>% mutate(resid_final = residuals(m_final))

ggplot(tracts) +
  geom_sf(aes(fill = resid_final), color = NA) +
  scale_fill_gradient2(low = "#2166ac", mid = "grey95", high = "#b2182b", midpoint = 0) +
  theme_void() +
  labs(title = "Final model: where it's wrong",
       subtitle = "Red = actual higher than predicted",
       fill = "Residual")
```

Put your two residual maps side by side, or put them next to each other in the rendered document. Did the clusters from Part 2C shrink? Where did they stay?

```{r}
#| label: residual-maps-comparison
#| fig-width: 12
#| fig-height: 5

map_data <- bind_rows(
  tracts %>%
    mutate(
      model = "Baseline model",
      residual = resid_m1
    ),
  tracts %>%
    mutate(
      model = "Final model",
      residual = resid_final
    )
)

ggplot(map_data) +
  geom_sf(aes(fill = residual), color = NA) +
  facet_wrap(~model, nrow = 1) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "grey95",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-50, 50)
  ) +
  theme_void() +
  labs(
    title = "Where the models are wrong",
    subtitle = "Red = actual higher than predicted · Blue = lower than predicted",
    fill = "Residual"
  )
```

The final model reduces some of the strongest residuals, with the maximum positive error falling from roughly 40 to 30 percentage points. Some northern and northeastern tracts become lighter, but errors remain spatially clustered around central, western, and southwestern Philadelphia. This suggests that the new spatial features helped, but the model still does not capture every local condition affecting household vehicle access.

### Are the biggest misses in the least reliable tracts?

```{r}
#| label: fog
ggplot(tracts, aes(x = y_cv, y = abs(resid_final))) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  labs(x = "CV of your outcome estimate (%)", y = "Size of the miss") +
  theme_minimal()
```

---

## Part 7: Reflection (about 300 words)

Answer in prose, drawing on your model log (incorporate our words of the day for bonus points. just kidding):

1. **What helped, and what didn't?** Name the feature that lowered CV RMSE the most and the one that disappointed you most. Offer a reason for each.
2. **Read one coefficient twice.** Choose a variable in your final model. Give its β̂ reading and its ŷ reading in one sentence each.
3. **What's still missing?** Look at your final residual map. Name one thing about those places you'd want to measure next, and where you might get the data.
4. **How much of the remaining error might be uncertainty or statistical noise?** Use your last plot.

In this analysis, I predicted the percentage of households without a vehicle across Philadelphia Census tracts. The squared City Hall distance helped the most, lowering CV RMSE from 12.46 in the baseline model to 9.76. This suggests that vehicle access changes with distance from the city center in a curved pattern rather than at one constant rate. The final model, which also included burglary density and distance to an Indego station, lowered RMSE slightly further to 9.71. The log transformation of burglary density disappointed me most. Although logging made the distribution appear more balanced, its RMSE was 11.64, slightly worse than the 11.53 produced by raw burglary density. A cleaner-looking distribution therefore did not automatically produce better predictions.

The estimated β̂ for renter percentage is 0.222, holding median income, educational attainment, burglary density, and the spatial variables constant. In ŷ terms, a tract with a renter share one percentage point higher is predicted to have about 0.22 percentage points more households without vehicles, assuming the other variables remain unchanged.

The final residual map shows smaller errors than the baseline map, but errors are still spatially clustered in parts of central, western, and southwestern Philadelphia. One missing feature may be the quality and frequency of public transportation. Indego stations measure bike-share proximity, but they do not represent subway, trolley, or bus service. I would obtain transit-stop locations and service-frequency information from SEPTA’s GTFS data and test whether better transit access explains the remaining clusters.

Finally, the uncertainty plot shows that prediction errors gradually increase at higher outcome CV values, although only a few tracts have extremely high CVs. Large errors also appear among tracts with relatively reliable estimates. Therefore, ACS uncertainty likely explains part of the remaining error, but the model is also missing neighborhood characteristics that influence whether households own vehicles.

---

## Submitting

Render this document to HTML and submit the link on Canvas by Friday, October 16. I'll miss you all terribly while I'm away. Please send letters or correspondences to my office mailbox. 

